Supplementary Information for\\ ``A weakly structured stem for human origins in Africa''

### 2023-03-11 supplement.tex
add some methods
add Brenna's supp section

\section{Data and sequencing}

@@

From the merged dataset of the African Diversity Reference Panel (ADRP)
\citep{Gurdasani2015-qy,Pagani2015-pz}
and the \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data,
we subsampled to the populations included in this study. From the 1000 Genomes
data, these included all populations in the African super-population (ESN, GWD,
LWK, MSL, and YRI) and four non-African populations (CEU, GBR, CHB, and PJL).
From the ADRP, these included the Amhara, Bakiga, Gumuz, Nama, Oromo, Somali,
Nama, and Zulu. This larger set of populations were used in the PCA and
\texttt{Admixture} analyses. The \texttt{Relate} analysis used a slightly
smaller set, excluding the Bakiga, Somali, and Zulu samples.

@@

\subsubsection{Fixing the East African and OOA divergence times}

In our preliminary models, the unconstrained inferred split times between eastern
African populations and Europeans was inferred to have occurred around $40$ka,
and the inferred split between eastern and western African populations occurred about
$10$ka earlier. These recent split dates are at odds with paleoanthropological 
evidence for an earlier Out-of-Africa (OOA) expansion
\citep[e.g.][]{Fu2014-am,Hublin2020-nk} and with the estimated date
for Neanderthal admixture into Eurasian populations
\citep[e.g.][]{Reilly2022-ym}. The parameter
uncertainty of our inferred divergence times (which accounts for the
non-independence of sites) does not account for these differences: in the
high-likelihood models, the $95\%$ confidence intervals for the eastern
African-European split was roughly $38 \pm 5$ka, and for the eastern/western African
split was $45 \pm 10$ka, with the uncertainties reflecting 2 standard
deviations according to bootstrap of genomic regions. We therefore expect these
discrepancies to result from model misspecification in the details of the
divergence between the ancestors of OOA and eastern African populations.

Our models suppose a single split leading to the OOA event followed by constant
symmetric migrations between eastern African and OOA populations, which is an
oversimplification. Previous studies that have focused on reconstructing the
OOA divergence time and subsequent migration rates typically included more
Eurasian populations and multiple epochs that allow for differing continuous
migration rates and population sizes. Since a detailed model for the OOA (and
Back-to-Africa) event was not our main focus, we chose to keep this simplified
model, but fix the East African-Eurasian split at $50$ka to be older than 
archaeological dates for when the OOA reached Europe 
\citep{Fewlass2020-se,Hublin2020-nk}. This also allowed us to include 
Neanderthal admixture into the OOA population at $45$ka.

Fixing these parameters (i.e., East Africa-Eurasian split at 50ka, Neanderthal
admixture at 45ka) resulted in an estimated value for the split between the
ancestors of East and West African populations at around $60$ka. This timing
agrees with previous LD-based estimates of West African-Eurasian divergence
times, of $\sim61$ka ($50-64$ka, $95\%$ confidence interval)
\citep{Ragsdale2019-nt}, which used a different set populations than those in
this study. A date of $60$ka is also in close agreement with inferred
divergence times of $\sim65$ka from both SFS-based demographic inference, using
synonymous variation from a subset of populations used in this study (Mende,
British, and Han Chinese), as well as that of LD-based inference using these
same groups, of $53$ka ($43-63$ka, $95\%$ confidence interval). Taken
altogether, we decided to fix this value ($60$ka) in most subsequent analyses.
We discuss the sensitivity of our model to these assumptions in
Section~\ref{sec:change-dates}.

@@

>>>>

\section{Data and sequencing}
\label{sec:data-sequencing}

@@

From the merged dataset of the African Diversity Reference Panel (ADRP)
\citep{Gurdasani2015-qy,Pagani2015-pz}
and the \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data,
we subsampled to the populations included in this study. From the 1000 Genomes
data, these included all populations in the African super-population (ESN, GWD,
LWK, MSL, and YRI) and four non-African populations (CEU, GBR, CHB, and PJL).
From the ADRP, these included the Amhara, Bakiga, Gumuz, Nama, Oromo, Somali,
and Zulu. This larger set of populations were used in the PCA and
\texttt{Admixture} analyses. The \texttt{Relate} analysis used a slightly
smaller set, excluding the Bakiga, Somali, and Zulu samples.

@@

\subsubsection{Evidence for the Out of Africa timing}

Above (Section~\ref{sec:models}), we assume that the eastern Africa-Eurasian
population divergence occurs at 50ka and Neanderthal admixture into this
Eurasian ancestral population occurs at 45ka. The choice of these dates
deserves additional discussion and consideration. Our focus is on the possible
expansion Out of Africa (OOA) that led to a sustained, widespread population
from whom present-day OOA populations are derived.

In this context, a number of possible models have been outlined
\citep{Groucutt2015-do}. Among these, \citeauthor{Groucutt2015-do} enumerate
variants on an early onset dispersal model. There is persuasive
paleo-morphological evidence for a Homo sapiens presence in Arabia between
120--85ka, coincident with Marine Isotope Stage (MIS) 5. The crania identified
at Skhul and Qafzeh in the Levant, as well a finger bone from southern Arabian
Peninsula point to a dispersal OOA when conditions in North Africa facilitated
a connection through the Sinai land bridge, and/or a possible connection via
crossing the Red Sea. Ecological conditions would have been similar between
North Africa and the Arabian Peninsula at this time, with greater abundance in
game during the higher rainfall MIS 5. While Skhul and Qafzeh are taxonomically
assigned to Homo sapiens, we note that their morphological diversity (i.e. a
mix of derived and archaic traits) is unique among contemporary human
populations. Under some formulations of the early onset dispersal model,
archaeologists posit a further expansion of humans into South Asia or as far as
southeastern Asia \citep{Groucutt2015-do}.

However, the replacement of Homo sapiens by Neanderthals in the Levant (e.g.
Kebara 2) complicates the persistence of modern humans in the Arabian Peninsula
after 85ka. The end of MIS 5 coincides with increasing aridity in the region
which would have rendered much the Peninsula uninhabitable, except for a
secluded southern refugia, decreasing the probability of migrations through the
Sinai land bridge \citep{Beyer2021-ym}. The northern route OOA remains a likely
barrier to human migration (as modeled by precipitation amount) until the
Holocene. A southern route across the Red Sea becomes possible during MIS 4
(70ka) until 30ka \citep{Beyer2021-ym}. Even during MIS 4, there were few
plausible routes out of the southern Arabian Peninsula, except along the Gulf
of Oman or gaps in the Arabian Desert through the Zagros Mountains. Neanderthal
remains dating to this period have been identified in the Zagros Mountains,
adding a further potential barrier to expansion
\citep{Yousefi2020-et,Heydari-Guran2021-xy}.

Thus, either an incipient OOA population must have expanded beyond the Near
East by 85ka and then given rise to today’s Eurasian populations or a
successful expansion would likely post-date 65ka. Morphological and
archaeological evidence for Homo sapiens in the Near East reappears by 50ka in
the Levant with the Initial Upper Paleolithic stone tool industry
\citep{Boaretto2021-oi} and a cranium from Manot Cave
\citep{Hershkovitz2015-lq}. There is also strong consensus that a widespread
human presence in Europe and Asia is dated to after 45ka. Archaeological
\citep[e.g.,][]{Fewlass2020-se}, morphological \citep[e.g.,][]{Hublin2020-nk},
and genetic data \citep[e.g.,][]{Prufer2021-ao} all indicate that about 45ka
\emph{Homo sapiens} had expanded into eastern Europe and within 5,000 years
rapidly colonized the entire span of southern Europe. This period further
coincides with the dates for the last Neanderthal presence in Europe at
approximately 40ka \citep{Deviese2017-de,Hublin2017-uy}. Similarly, by 45ka, an
unequivocal \emph{H. sapiens} femur from southern Siberia contains derived
alleles shared by later Eurasian populations \citep{Fu2014-am}. The Ust’-Ishim
specimen contains evidence of admixture with Neanderthals, at a level similar
to contemporary East Asian genomes. This supports estimates of Neanderthal
admixture occurring earlier than 45ka, estimated to be by 47-65ka as inferred
from Neanderthal tract lengths \citep{Sankararaman2012-uw}.

In summary, a sustained, rapid human expansion across Eurasia dates to just
prior to 45ka and emerges from a Near Eastern population. The time of
population divergence between eastern Africans and this Near Eastern group is
difficult to pinpoint. An exit OOA and persistence prior to 65ka appears
unlikely to have contributed to the ancestors of most Eurasians. By 50ka, a
Homo sapiens population occupied the Levant, subsequently admixed and replaced
Neanderthals, and is a likely source for the Eurasian populations. Hence, we
fixed the eastern African/Eurasian divergence at 50ka, though arguably dates
even earlier up to 65ka would be appropriate (see SI Section 7.2).

We tested our models for sensitivity to a subset of the fixed parameters,
including those relevant to the OOA timing. When unconstrained, the inferred
population split times between eastern African populations and Europeans
occurred around 40ka, and the inferred split between eastern and western
African populations occurred about 10ka earlier. The 95\% CI for the inferred
divergence times (which accounts for the non-independence of sites) does not
contain the values of the fixed estimates. Under the high-likelihood models,
the 95\% confidence intervals for the eastern African-European divergence were
roughly $38\pm5$ka, and for the eastern/western African divergence was
$45\pm10$ka, with the uncertainties reflecting 2 standard deviations according
to bootstrap of genomic regions.

We hypothesize that these discrepancies result from model misspecification
related to the divergence between the ancestors of OOA and eastern African
populations. As is typical, our models suppose a single split leading to the
OOA event followed by constant, symmetric migration between eastern African and
OOA populations until 12ka. We consider the British population (GBR) as a proxy
for all Eurasians populations, in part to avoid adding many additional
parameters associated with including 2 or 3 Eurasian populations. Further, our
study and most others lack representative genomes from the Arabian Peninsula
and Levant.

@@

In summary, fixing these parameters (i.e., East Africa-Eurasian split at 50ka,
Neanderthal admixture at 45ka, see Section~\ref{sec:models}) resulted in an
estimated value for the split between the ancestors of East and West African
populations at around $60$ka. This timing agrees with previous LD-based
estimates of West African-Eurasian divergence times, of $\sim61$ka ($50-64$ka,
$95\%$ confidence interval) \citep{Ragsdale2019-nt}, which used a different set
populations than those in this study. A date of $60$ka is also in close
agreement with inferred divergence times of $\sim65$ka from both SFS-based
demographic inference, using synonymous variation from a subset of populations
used in this study (Mende, British, and Han Chinese), as well as that of
LD-based inference using these same groups, of $53$ka ($43-63$ka, $95\%$
confidence interval). Taken altogether, we decided to fix this value ($60$ka)
in most subsequent analyses.
